Syntax Is All You Need: A Universal-Language\\Approach to Mutant Generation

### 2023-03-28 paper/main.tex
Adding URLs

%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Syntax Is All You Need: A Universal-Language Approach to Mutant Generation}

@@

\begin{abstract}
While mutation testing has been a topic of academic interest for
decades, it is only recently that ``real-world'' developers, including
industry leaders such as Google and Meta, have adopted mutation
testing.  In this paper we propose a new approach to the development of mutation
testing tools, and in particular the core challenge of
\emph{generating mutants}.  Current practice tends towards two
limited approaches to mutation generation: mutants are either (1)
generated at the bytecode/IR level, and thus neither human readable
nor adaptable to source-level features of languages or projects, or
(2) generated at the source level by language-specific tools that are
hard to write and maintain, and in fact are often abandoned by both
developers and users.  We propose instead that source-level mutation
generation is a special case of \emph{program transformation} in
general, and that adopting this approach allows for a single tool that
can effectively generate source-level mutants for essentially
\emph{any} programming language. Furthermore, by using \emph{parser
  parser combinators} many of the seeming limitations of an
any-language approach can be overcome, without the need to parse
specific languages.  We compare the universal program transformation
approach to mutation to existing tools, and demonstrate the advantages
of using parser parser combinators to improve on a simple regular-expression
based approach to generation.  
\end{abstract}

@@

However, the same website demonstrates the existence of a sad history
of mutation testing tools: mutation tools have a tendency to be
developed for a while and then abandoned.   The page listing abandoned
mutation testing tools
(\url{https://github.com/theofidry/awesome-mutation-testing/blob/master/abandoned.md})
lists almost half as many tools as are under active maintenance, and
these tools cover approximately half as many languages.  Each of these
tools required considerable development time and effort, and was
popular enough at some point to be worth noting in the ``Awesome
Mutation Testing'' resource.  Many of the abandoned tools have GitHub
repositories (some archived) with more than 20 stars, and so may well
have had users who at some point faced the reality of no future bug
fixes or changes to address changes to the programming language being
mutated\footnote{While this is a minor concern for languages such as C
  that are relatively stable, it means that a mutation tool may fail
  to parse C++ 2020, and is particularly problematic for somewhat
  fast-moving languages such as Rust.}.  There
appears to be a significant risk in relying on any mutation tools other
than the most widely used and well-supported, such as PIT.

Moreover, newer languages, including ones with considerable interest
in the development community, often lack mutation testing tools.  Even
languages with a long history may long lack a working mutation testing
tool; e.g., it was not until 2014 that Haskell had such a tool
\cite{mucheck}.  Furthermore, that tool is no longer
supported\footnote{The last commits or package updates were in 2015, and the
primary author confirms that changes to {\tt haskell-src-ext} since
introduced render it unusable at present.}, and it
appears that no successor tool has appeared in the ensuing
years\footnote{There is FitSpec \cite{FitSpec}, which is maintained,
  but FitSpec does not perform traditional mutation testing or produce
  source code (or bytecode) mutants; it rather
  performs black-box type-aware mutation of function return values.}.
Haskell developers (until now, see below) have no option for
performing true mutation testing.

@@

\begin{enumerate}
  \item A single tool can provide mutant generation for almost any
    programming language, without needing to parse that language, and
    rules for one language can be re-used for another language.  This
    reduces the development and maintenance effort for mutation
    testing tools by orders of magnitude.  Moreover, it removes the
    need for many   maintenance activities related to parsing code;
    such changes need to be made only when a change to the language
    would introduce new mutation operators or modify existing ones.
    The local nature of operators tends to make the second case very rare.
    \item The existing literature on multi-lingual or language
      agnostic program transformation can be applied to the problem of
      mutant generation, making it easy to express mutation operators
      that operate across many languages, and efficient to generate
      valid mutants even in the absence of language-specific
      development effort.
      \end{enumerate}

@@

          \item A proposal to use parser parser combinators to improve the
      efficiency and expressive power of that approach, and an
      evaluation of the gains thus achieved.

@@

\begin{table}[hbtp]
\centering
\caption{Rule Examples}
\label{tab:rules}
\resizebox{\columnwidth}{!}{%

  \begin{tabular}{|l|l|l|}
    \hline
    Universal & C-Like & Language-Specific \\
    \hline
{\tt < ==> >} &{\tt  else ==>} & {\tt synchronized ==>} \\ 
{\tt < ==> ==} & {\tt if (\textbackslash(.*\textbackslash)) ==> if (!\textbackslash 1)} &
                                                                 {\tt (\^{}\textbackslash s*)(\textbackslash S+.*\textbackslash n) ==> \textbackslash 1pass}\\
{\tt < ==> <=} & {\tt if (\textbackslash(.*\textbackslash)) ==> if
                 (1==1)} & {\tt any\_of ==> all\_of}\\
\hline

  Table~\ref{tab:rules} shows examples of universal, C-like, and
  language-specific rules for a few languages, taken directly from our
  implementation.  In all cases, the form of a rule is: {\tt regexp
    ==> replacement}, using Python regular expression and replacement syntax.

  The first column shows universal rules applied to
  all languages, here drawn from the large set of transformations for
  less-than comparisons.  The second column shows somewhat more
  complex rules, for removing elses (which always leaves valid code in
  C-like languages, but with the else clause always executing) and for
  replacing if conditions with, respectively, a negation and a
  constantly true expression.  Finally, the last column shows
  language-specific rules, for, respectively, Java, Python (statement
  deletion, including a {\tt pass} and respecting indentation level),
  and C++.  The last rule was contributed by a user using our tool to
  perform mutation testing on the bitcoin core implementation.  For a
  C++ program, all of the rules in the first two columns will be
  applied, as well as the last rule in the third column.  For a Python
  program, however, only the first column of rules and the second rule
  in the last column apply, as Python is not sufficiently ``C-like''.

@@

We implemented our approach in a mutation testing tool, written in
about 2,200 lines of Python.  The rules defining mutation operators
for a set of languages including C, C++, Python, Java, Lisp, Swift,
Rust, Solidity, Fe, Vyper, JavaScript, and Go, required another 436
lines; this is the total for comby and regexp-based rules, so each
version requires only 218 lines of rule definitions.

While many tools provide a GUI front-end, or integrate to a particular
testing library, we chose to focus on an easily configured CLI,
allowing a user to specify the commmands used to build mutants or
run tests.  While less convenient in the case where there is a good
fit between a tool's integrations and a project's environment, we
found this flexibility useful even when limiting considerations to a
single language.  For example, the widely used PIT system is strongly tied to
certain versions of JUnit, and mutating projects using a different
testing infrastructure, or an older JUnit, can be difficult.  Our
approach also is suitable for use in CI and other automated environments.

For the most part we expect users to provide build commands, but do
support automatic compilation and comparison of mutants for Python,
Solidity, Fe, Vyper, Swift, Rust, and Java, including Trivial Compiler
Equivalence \cite{TCE} checks for redundancy of mutants.  These
default handlers can always be over-ridden if they do not work with a
particular build setup.

Our tool has been available as an open source project on GitHub and
available for installation via Python pip for
some time, has received contributions from users not involved in
our project, and been adopted by at least one very large project as
the officially recommended mutation tool.  We omit details for
purposes of blinding.

@@

  \item{RQ1:}  How does the use of parser parser combinators modify 
    the efficiency (in terms of invalid mutants) and effectiveness (in
    terms of mutation score and equivalent mutants) of the universal
    approach to mutant generation?

\item{RQ2:}  How does the univeral approach, using regular expressions
  or parser parser combinator rules, compare to existing
  single-language mutation tools, in terms of number of generated
  mutants, mutation scores, equivalent mutants, and other evaluation measures used in the literature?

@@

Both research questions were addressed by generating and running
mutants for a set of four programming languages: C++, Java, Python,
and Rust.  We picked six projects with 1) a large number of GitHub
stars and 2) an easy-to-execute, passing test suite for each language,
and then chose the largest source file below 20KB in size from each project to mutate.  Limiting
mutation to one large but not extremely large file made is easier to compare generated mutants and
manually inspect for mutant equivalency.  The only bias to file
selection was size, which we do not believe should affect our results
significantly.

We additionally present a small
case study, applying our implementation to a program in a language for
which it has \emph{no} rules, Haskell.

@@

@@

\begin{table}[hbtp]
  \centering
  \caption{Overview of Mutation Testing Tools.}
  \label{tab:mutationtools}

  \begin{tabular}{|c|c|r|}
  \hline
  \textbf{Language} & \textbf{Tool} & \textbf{LOC}  \\
  \hline
  \multirow{2}{*}{Python}  & Mutmut &	3870  \\\cline{2-3}
      & cosmic-ray & 4599 \\ \hline
  \multirow{2}{*}{Java}  & PIT &	59577  \\\cline{2-3}
      & LittleDarwin & 22359 \\ \hline
      Rust & Cargo-Mutants & 7020 \\\hline
      C++ & Dextool & 38611 \\\hline
      ALL & universalmutator & 2680 \\\hline
  \end{tabular}
\end{table}

Table \ref{tab:mutationtools} shows the set of tools compared.  We
attempted to identify well-maintained and commonly used tools for each
language chosen.

@@

\begin{figure}

  \begin{code}
 qsort :: [ Int] -> [ Int ]
 qsort [] = []
 qsort (x: xs ) = qsort l ++ [x] ++ qsort r
     where l = filter (< x) xs
           r = filter ( >= x) xs
\end{code}
\caption{Example from MuCheck Paper}
\label{fig:haskell}

\end{figure}

@@

Using regular expressions, and only the universal rules, our tool
generates 10 valid mutants and 34 invalid mutants of the code (22.73\%
valid mutants).  Comby produces an identical 10 valid mutants, but 65
invalid mutants (13.33\% validity).  Both tools produce a large number
of invalid mutants here due to including rules such as adding {\tt
  break} and {\tt continue} that are sufficiently widespread in
applicability to be included in our ``universal'' set but do not apply
to Haskell; we believe that comby's awareness of Haskell syntax in
this case actually gives it \emph{more} opportunities to apply such
inapplicable mutants.

These mutants produce a mutation
score of 0.8 for the two properties defined in the original MuCheck
paper.  In other words, like MuCheck, our tool is able to
determine that the properties provided are insufficient.  In
particular, while they check idempotence and that the result is
sorted, they do not check that the result of {\tt qsort} is a permutaton of the input
list.  The MuCheck paper only provides mutation scores for the two
properties independently.  For idempotence, MuCheck gives a mutation
score of 0.84, and our implementation yields a score of 0.8.  Checking
only sortedness, MuCheck yields a score of 0.61, while our approach
produces an even lower score of 0.4.  Our lower
scores are not due to equivalent mutants: adding a property to check
that the sorted list is a permutation of the original list, all
mutants generated by our approach can be killed.

@@

The approach reported in this paper was briefly presented in part in a
tool paper (citation redacted for blinding purposes) that introduced
the idea of using regular expressions to generate mutants for multiple
languages, and provided an evaluation against PIT for a single toy
Java program.  The present paper provides the full idea and context,
provides empirical data for universal mutant generation on real world
programs over four languages,
and introduces the idea of using parser parser combinators to improve
on regular expression-based generation.

>>>>

%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Syntax Is All You Need: A Universal-Language\\Approach to Mutant Generation}

@@

\begin{abstract}
While mutation testing has been a topic of academic interest for
decades, it is only recently that ``real-world'' developers, including
industry leaders such as Google and Meta, have adopted mutation
testing.  In this paper we propose a new approach to the development of mutation
testing tools, and in particular the core challenge of
\emph{generating mutants}.  Current practice tends towards two
limited approaches to mutation generation: mutants are either (1)
generated at the bytecode/IR level, and thus neither human readable
nor adaptable to source-level features of languages or projects, or
(2) generated at the source level by language-specific tools that are
hard to write and maintain, and in fact are often abandoned by both
developers and users.  We propose instead that source-level mutation
generation is a special case of \emph{program transformation} in
general, and that adopting this approach allows for a single tool that
can effectively generate source-level mutants for essentially
\emph{any} programming language. Furthermore, by using \emph{parser
  parser combinators} many of the seeming limitations of an
any-language approach can be overcome, without the need to parse
specific languages.  We compare the universal program transformation
approach to mutation to existing tools, and demonstrate the advantages
of using parser-parser combinators to improve on a simple regular-expression
based approach to generation.  
\end{abstract}

@@

However, the same website demonstrates the existence of a sad history
of mutation testing tools: mutation tools have a tendency to be
developed for a while and then abandoned.   The page listing abandoned
mutation testing tools
(\url{https://github.com/theofidry/awesome-mutation-testing/blob/master/abandoned.md})
lists almost half as many tools as are under active maintenance, and
these tools cover approximately half as many languages.  Each of these
tools required considerable development time and effort, and was
popular enough at some point to be worth noting in the ``Awesome
Mutation Testing'' resource.  Many of the abandoned tools have GitHub
repositories (some archived) with more than 20 stars, and so may well
have had users who at some point faced the reality of no future bug
fixes or changes to address changes to the programming language being
mutated\footnote{While this is a minor concern for languages such as C
  that are relatively stable, it means that a mutation tool may fail
  to parse C++ 2020, and is particularly problematic for somewhat
  fast-moving languages such as Rust.}.  Of course, many open source
projects are abandoned, and academic software has a long history of
being available and supported only so long as a graduate student is
working on a thesis or a particular grant is being funded.  Still,
there appears to be a significant risk in relying on any mutation tools other
than the most widely used and well-supported, such as PIT.

Moreover, newer languages, including ones with considerable interest
in the development community, often lack mutation testing tools.  Even
languages with a long history may long lack a working mutation testing
tool; e.g., it was not until 2014 that Haskell had such a tool
\cite{mucheck}.  Furthermore, that tool may no longer  be
supported (the last commits or package updates were in 2015), and it
is not clear that a successor tool has appeared in the ensuing years.

@@

\begin{enumerate}
  \item A single tool can provide mutant generation for almost any
    programming language, without needing to parse that language, and
    rules for one language can be re-used for another language.  This
    reduces the development and maintenance effort for mutation
    testing tools by orders of magnitude.  Moreover, it removes the
    need for many   maintenance activities related to parsing code;
    such changes need to be made only when a change to the language
    would introduce new mutation operators or modify existing ones.
    The local nature of operators tends to make the second case in
    particular very rare.
    \item The existing literature on multi-lingual or language
      agnostic program transformation can be applied to the problem of
      mutant generation, making it easy to express mutation operators
      that operate across many languages, and efficient to generate
      valid mutants even in the absence of language-specific
      development effort.
      \end{enumerate}

@@

          \item A proposal to use parser-parser combinators to improve the
      efficiency and expressive power of that approach, and an
      evaluation of the gains thus achieved.

@@

  Figure~\ref{fig:rules} shows examples of universal, C-like, and
  language specific rules for a few languages, taken directly from our
  implementation.

@@

Our approach also makes introducing advanced techniques such as
Trivial Compiler Equivalence (TCE) into a mutation workflow easy.
While we support TCE for some languages inside the tool, a user can
always write a simple script to check...

@@

  \item{RQ1:}  How does the use of parser-parser combinators modify 
    the efficiency (in terms of invalid mutants) and effectiveness (in
    terms of mutation score and equivalent mutants) of the universal
    approach to mutant generation?

\item{RQ2:}  How does the univeral approach, using regular expressions
  or parser-combinator combinator rules, compare to existing
  single-language mutation tools, in terms of number of generated
  mutants, mutation scores, equivalent mutants, and other evaluation measures used in the literature?

@@

In addition to these core research questions, evaluated over six files
from six real-world projects for four languages, we also present a
case study, applying our implementation to a program in a language for
which it has no rules, Haskell.

@@

\begin{table*}[hbtp]
  \centering
  \caption{Overview of Mutation Testing Tools. universalmutator has singificantly fewer lines of code than any of the single language tools.}
  \label{tab:mutationtools}

  \begin{tabular}{|c|c|c|r|}
  \hline
  \textbf{Language} & \textbf{Tool} & \textbf{URL} & \textbf{LOC}  \\
  \hline
  \multirow{2}{*}{Python}  & Mutmut & https://github.com/boxed/mutmut & 3870  \\\cline{2-4}
      & cosmic-ray & https://github.com/sixty-north/cosmic-ray & 4599 \\ \hline
  \multirow{2}{*}{Java}  & PIT & https://github.com/hcoles/pitest & 59577  \\\cline{2-4}
      & LittleDarwin & https://github.com/aliparsai/LittleDarwin & 22359 \\ \hline
      Rust & Cargo-Mutants & https://github.com/sourcefrog/cargo-mutants & 7020 \\\hline
      C++ & Dextool & https://github.com/joakim-brannstrom/dextool & 38611 \\\hline
      ALL & universalmutator & https://github.com/agroce/universalmutator & 2244 \\\hline
  \end{tabular}
  \end{table*}

@@

@@

\begin{figure}
\begin{code}
 qsort :: [ Int] -> [ Int ]
 qsort [] = []
 qsort (x: xs ) = qsort l ++ [x] ++ qsort r
     where l = filter (< x) xs
           r = filter ( >= x) xs
\end{code}
\caption{Example from MuCheck Paper}
\label{fig;haskell}
\end{figure}

@@

Using regular expressions, and only the universal rules, our tool
generates 10 valid mutants and 34 invalid mutants of the code (22.73\%
valid mutants).  Comby produces an identical 10 valid mutants, but 65
invalid mutants (13.33\% validity).  These mutants produce a mutation
score of 0.8 for the two properties defined in the original MuCheck
paper.  In other words, like MuCheck, our tool is able to
determine that the properties provided are insufficient.  In
particular, while they check idempotence and that the result is
sorted, they do not check that the result of {\tt qsort} is a permutaton of the input
list.  The MuCheck paper only provides mutation scores for the two
properties independently.  For idempotence, MuCheck gives a mutation
score of 0.84, and our implementation yields a score of 0.8.  Checking
only sortedness, MuCheck yields a score of 0.61, while our approach
produces an even lower score of 0.4.  Our lower
scores are not due to equivalent mutants: adding a property to check
that the sorted list is a permutation of the original list, all
mutants generated by our approach can be killed.

@@

The approach reported in this paper was briefly presented in part in a
tool paper (citation redacted for blinding purposes) that introduced
the idea of using regular expressions to generate mutants for multiple
languages, and provided an evaluation against PIT for a single toy
Java program.  The present paper provides the full idea and context,
provides empirical data for universal mutant generation on real world
programs over four languages,
and introduces the idea of using parser-parser combinators to improve
on regular expression-based generation.
